---
title: "COMP2501 In-class Exercises 01"
output:
  html_document:
    df_print: paged
  word_document: default
---

<!-- Knit this in RStudio or posit Cloud. Generated from the matching .ipynb. -->

# COMP2501 · In-class Exercises 01 (interactive)

This is the official **In-class Exercises 01** from the course, rebuilt as an interactive walkthrough. The goal is the same as in class: **try different code, predict outputs, and understand the dplyr verbs** — not just copy-paste.

**What you need:** `dplyr`, `ggplot2`, `datasets`. If a package is missing, run the install cell below (once).

> 🎓 **Exam-relevant.** `mutate`, `filter`, `select`, `group_by`/`summarize`, `arrange`, `top_n` are the core dplyr verbs and come up repeatedly. Get comfortable with the *pipe* `|>` too — it's used throughout.

---
## 0 · Packages

Load the library, install once if needed.

```{r}
# install.packages(c("dplyr", "ggplot2"))   # run once if not installed

library(dplyr)
library(ggplot2)
library(datasets)
cat("packages loaded OK\n")
```

---
## 1 · Write a function

Write `compute_s_n(n)` that returns the sum of squares:

$$S_n = 1^2 + 2^2 + 3^2 + \dots + n^2$$

Then report the value for `n = 100`.

```{r}
# YOUR CODE
# compute_s_n <- function(n) { ... }
# compute_s_n(100)
```

<details>
<summary><b>Reveal answer</b></summary>

```r
compute_s_n <- function(n) {
  s <- 0
  for (i in 1:n) {
    s <- s + i^2
  }
  return(s)
}

compute_s_n(100)   # 338350
```

The for-loop version works because R's `1:n` makes a sequence. Note: `s <- s + i^2` accumulates. (vectorized alt: `sum((1:100)^2)`)
</details>

---
## 2 · Explore `mtcars` with dplyr

Load the built-in dataset and inspect it.

```{r}
data("mtcars")
head(mtcars)      # first 6 rows
dim(mtcars)       # observations x variables
names(mtcars)     # all variable names
```

**b) Create `performance = hp / wt`** on a copy called `mtcars_modified`. Use `mutate`.

```{r}
# YOUR CODE
# mtcars_modified <- mtcars |> mutate(performance = ?)
# head(mtcars_modified)
```

<details>
<summary><b>Reveal answer</b></summary>

```r
mtcars_modified <- mtcars |> mutate(performance = hp / wt)
head(mtcars_modified)
```

`mutate` adds/overwrites a column. The pipe `|>` feeds `mtcars` into `mutate` as its first arg. `hp / wt` is vectorized (each row divided).
</details>

**c) 1) Top-1 by `performance`; 2) sort descending.** Use `top_n` and `arrange`.

```{r}
# YOUR CODE
# 1. top_n(mtcars_modified, 1, performance)
# 2. mtcars_modified |> arrange(desc(performance)) |> head()
#    is the top_n result the first row of the sorted result?
```

<details>
<summary><b>Reveal answer</b></summary>

```r
top_n(mtcars_modified, 1, performance)
mtcars_modified |> arrange(desc(performance)) |> head()
```

Yes — the `top_n` row is the first row of the `arrange(desc(...))` result. `top_n(1)` = the single highest; `arrange(desc(perf))` = full data sorted descending, so its first row matches.
</details>

**d) Filter to mpg > 20** → `mtcars_filtered` (use `filter`).

```{r}
# YOUR CODE
# mtcars_filtered <- mtcars |> filter(?)
# head(mtcars_filtered)
```

<details>
<summary><b>Reveal answer</b></summary>

```r
mtcars_filtered <- mtcars |> filter(mpg > 20)
head(mtcars_filtered)
```

`filter` keeps rows where the condition is TRUE. Note this filters the whole row, not columns — that's `select`'s job.
</details>

**e) Select only `mpg` and `cyl`** from the filtered set → `mtcars_selected` (use `select`).

```{r}
# YOUR CODE
# mtcars_selected <- mtcars_filtered |> select(?)
# head(mtcars_selected)
```

<details>
<summary><b>Reveal answer</b></summary>

```r
mtcars_selected <- mtcars_filtered |> select(mpg, cyl)
head(mtcars_selected)
```

`select` picks **columns** (keeps all rows). `filter` picks **rows** (keeps all columns). Easy to mix up — remember, filter = rows, select = columns.
</details>

**f) Group by `cyl`, summarize mean mpg** → `mtcars_grouped` (use `group_by` + `summarize`).

```{r}
# YOUR CODE
# mtcars_grouped <- mtcars_selected |> group_by(cyl) |> summarize(?)
# mtcars_grouped
```

<details>
<summary><b>Reveal answer</b></summary>

```r
mtcars_grouped <- mtcars_selected |> group_by(cyl) |> summarize(mean_mpg = mean(mpg))
mtcars_grouped
```

`group_by(cyl)` splits data by cylinder value; `summarize(mean_mpg = mean(mpg))` aggregates each group to one row. Result: one mean mpg per cylinder group. This is the classic **split-apply-combine** pattern — the heart of data analysis.
</details>

---
## Recap — the 5 dplyr verbs + pipe

| verb | does | rows/cols |
|------|------|-----------|
| `filter` | keep rows matching a condition | rows |
| `select` | keep chosen columns | cols |
| `mutate` | add / overwrite a column | cols |
| `arrange` | sort rows | rows |
| `summarize` (with `group_by`) | collapse groups to summaries | rows ↓ |

The pipe `|>` threads the left side as the first argument to the right side:
`df |> filter(...) |> select(...)` = `select(filter(df, ...), ...)`. It makes the chain read top-to-bottom like a recipe.

**Quick self-quiz:**  
- `filter` vs `select`: which drops columns, which drops rows?
- What does `mutate` do that `summarize` doesn't?
- What does `group_by` *alone* do to a data frame? (It doesn't change the rows you see — it just remembers the grouping for the next verb.)

Next up: `lec02-tidyverse` for importing data and more dplyr.
