---
title: "COMP2501 Lec 1 - R Basics"
output:
  html_document:
    df_print: paged
  word_document: default
---

<!-- Knit this in RStudio or posit Cloud. Generated from the matching .ipynb. -->

# COMP2501 · Lec 1 — R Basics

**Course**: Introduction to Data Science and Engineering (RB Luo)  
**Goal**: the very basics of R — objects, assignment, functions, atomic types, and data frames.

> **How to use this notebook** (the loop):
> 1. Read the concept cell. **Predict** what the output will be, out loud or in your head.
> 2. Write the code into the **blank** cell.
> 3. Run it. Compare against your prediction.
> 4. Only then reveal the **Answer** cell.
>
> This is not about writing R fast — it's about *understanding what the code does*. The course lets you use AI to write code, but you must be able to read and predict the output yourself.

---
## Before you start: set the runtime to R

This notebook needs the **R** runtime. In Colab:
`Runtime ▸ Change runtime type ▸ R`  
(If you opened this from GitHub, Colab may already detect R and offer to switch.)

---
## The big picture: why R

The lecture makes a few key points — read them, they'll save you confusion later:

- **R is not a "software engineering" language.** It was built by *statisticians* as an interactive environment for data analysis. It won't follow the conventions you know from C/Java/Python — don't fight it.
- **Interactivity is the point.** You explore data quickly. Hence the console, not just scripts.
- **You can save your work as scripts** — the key feature that makes analysis *reproducible*.
- **R's superpower is visualization**, ahead of Python (pandas). We'll get there.

**The running joke in the slides:** R looks easy to learn (`x <- 1`) but subtle to master (integer vs numeric, vectorization, recycling). This notebook only covers the easy half — but the *why* behind it.

---
## 1 · Objects & assignment

In R, a variable is called an **object**. You assign with `<-` (a left arrow), **not** `=`.

```r
a <- 1   # assign 1 to object 'a'
b <- 1
c <- -1
```

This solves the quadratic formula example from the lecture: `x^2 + x - 1 = 0` → `x = (-b ± √(b²-4ac)) / 2a`.

**Why `<-` and not `=`?**  
- `=` also works in R, but is **not recommended** — it first means "give" in some contexts and can confuse.
- `<-` reads as **"assign to"** and makes intent unambiguous.
- When you see `x == 1`, that's **comparison** (equal). `x <- 1` is **assignment**. Different arrows, different meaning.

> 💡 **Snake_case**: the lecture recommends `solution_1` over `s1`, and describes R names with `all_lower_case` + `underscores`. Be descriptive.

**Task 1 — solve the quadratic.** Write `a`, `b`, `c` for `x^2 + x - 1 = 0`. Then compute the two roots using `sqrt()` and the quadratic formula.

```{r}
# YOUR CODE
# 1. assign a, b, c
# 2. solution_1 <- (-b + sqrt(b^2 - 4*a*c)) / (2*a)
# 3. solution_2 <- (-b - sqrt(b^2 - 4*a*c)) / (2*a)
# 4. print both
```

<details>
<summary><b>Reveal answer</b></summary>

```r
a <- 1
b <- 1
c <- -1

solution_1 <- (-b + sqrt(b^2 - 4*a*c)) / (2*a)
solution_2 <- (-b - sqrt(b^2 - 4*a*c)) / (2*a)

solution_1
solution_2
```

You should get approximately `0.618` and `-1.618` (the golden-ratio pair).
</details>

---
## 2 · The workspace

Every object you create lives in the **workspace**. Inspect it with:

```{r}
# YOUR CODE
# print all objects currently in your workspace (hint: ls)
```

<details>
<summary><b>Reveal answer</b></summary>

```r
ls()
```

Note the parentheses — `ls` is a **function**, and you must *call* it with `()`. Typing `ls` without `()` just shows the function definition, not its result. Try it.
</details>

### Remove an object, and a subtlety: case-sensitivity

- `rm(a)` removes object `a`. After that, `a` gives an error: `object 'a' not found`.
- **R is case-sensitive.** `True` (uppercase T) is **not** the logical value — it's an undefined object → error. The logical values in R are `TRUE` and `FALSE` (or `T`/`F`).

**Try this yourself:** remove one of your objects, then reference it. See the error.

```{r}
# YOUR CODE
# 1. remove 'solution_1' using rm()
# 2. print solution_1 (expect an error)
# 3. try: True  (expect an error, not TRUE)
# 4. print the logical TRUE
```

<details>
<summary><b>Reveal answer</b></summary>

```r
rm(solution_1)
solution_1        # Error: object 'solution_1' not found

True              # Error: object 'True' not found (case matters!)
TRUE              # [1] TRUE
```
</details>

---
## 3 · Functions

R is a **functional** language — almost everything is a function call. Two things matter most:

**a) Arguments, and their defaults.** Look at `log`:

```r
log(10)        # [1] 2.302585  (natural log, base e)
log(8, 2)      # [1] 3        (base 2)
log(x = 8, base = 2)
log(base = 2, x = 8)   # named args → order doesn't matter
```

The help page shows `log(x, base = exp(1))`. So `x` is **required**; `base` is **optional** with a default. Named arguments let you skip the order.

**b) Looking things up.** `?log` (or `help("log")`) opens the docs. Try `?"+"` and `?"^"` for operators. (Bonus: `?l"og without parens` shows how typing a built-in function name alone just echoes its definition.)

**Task 2 — predict, then confirm.** Before running, predict the output of each line:

```{r}
log(100)          # predict: ??
log(100, 10)     # predict: ??
log(2, 8)        # predict: ??  (careful — is this log_2(8) or log_8(2)?)
log(base = 2, x = 8)
```

<details>
<summary><b>Reveal answer</b></summary>

```r
log(100)          # [1] 4.60517  (natural log)
log(100, 10)     # [1] 2        (log base 10)
log(2, 8)        # [1] 0.3333   (this is log_8(2) = 1/3 — arg order = f(x, base))
log(base = 2, x = 8)   # [1] 3   (now x=8, base=2 → log_2(8)=3)
```

The third line is the classic gotcha: **positional arguments** go in the order `(x, base)`, so `log(2, 8)` means log with x=2, base=8. If you meant log base 2 of 8, you must either spell out `log(base=2, x=8)` or use `log(8, 2)`.
</details>

---
## 4 · Atomic data types

Use `class()` to see an object's type. The major atomic types in R:

| type | meaning | example | `class()` result |
|------|---------|---------|------------------|
| `numeric` | float/double | `3` | `"numeric"` |
| `integer` | whole number | `3L` (note the L) | `"integer"` |
| `logical` | boolean | `TRUE` | `"logical"` |
| `character` | string | `"3"` | `"character"` |
| `factor` | categorical | (later) | `"factor"` |

**Task 3 — classify.** Predict each `class()` result, then run:

```{r}
class(3)
class(3L)
class(TRUE)
class("3")
class(3.0)

# and the .is. family:
is.numeric(3)
is.integer(3)
is.integer(3L)
is.character("3")
is.logical(TRUE)
```

<details>
<summary><b>Reveal answer</b></summary>

```r
class(3)       # [1] "numeric"
class(3L)      # [1] "integer"
class(TRUE)    # [1] "logical"
class("3")    # [1] "character"
class(3.0)     # [1] "numeric"   <- even "3.0" is a double!

is.numeric(3)      # TRUE
is.integer(3)      # FALSE  (3 is a double, not integer — the lecture's subtle point)
is.integer(3L)     # TRUE
is.character("3") # TRUE
is.logical(TRUE)   # TRUE
```

The key insight: `3` and `3L` are **different** types. `3` is a `numeric`/double; `3L` is `integer`. R auto-coerces in most cases, but the difference matters for: indexing, memory, and why `is.integer(3)` is `FALSE`.
</details>

### Integer vs numeric, and indexing

The lecture's punchline: **integer is mostly used for indexing.**

```r
c(1, 2, 3, 4)[1.99] == c(1, 2, 3, 4)[1]   # TRUE — R truncates the index
```

So `c(...)` builds a vector (combine elements). `vec[1]` grabs the first element. R is **1-indexed** (starts at 1, not 0 like Python). **Try it.**

```{r}
# YOUR CODE
# 1. build a vector of 1,2,3,4 with c()
# 2. grab its 2nd element with [2]
# 3. check: c(1,2,3,4)[1.99] == c(1,2,3,4)[1]
# 4. what index is the first element in R? (1 or 0?)
```

<details>
<summary><b>Reveal answer</b></summary>

```r
v <- c(1, 2, 3, 4)
v[2]                        # [1] 2
c(1, 2, 3, 4)[1.99] == c(1, 2, 3, 4)[1]   # TRUE
# first element is index 1 (R is 1-indexed, unlike Python's 0-indexed)
```
</details>

---
## 5 · Data frames

A single number isn't useful for real data. The workhorse of R is the **data frame**: a table with rows (observations) and columns (variables), where each column can be a different type.

The lecture uses the `murders` dataset (from the `dslabs` package) and base datasets. We'll use a built-in base dataset first, then optionally the murders one.

**Three functions to inspect any data frame:**
- `head(df)` / `tail(df)` — first/last 6 rows
- `str(df)` — structure: how many obs/vars, and the type of each column
- `names(df)` — the column names

```{r}
# YOUR CODE — inspect the built-in 'mtcars'
# 1. load it: data("mtcars")
# 2. head(mtcars)
# 3. str(mtcars)
# 4. names(mtcars)
# 5. dim(mtcars)  (rows, columns)
# 6. how many rows? how many columns?
```

<details>
<summary><b>Reveal answer</b></summary>

```r
data("mtcars")
head(mtcars)
str(mtcars)
names(mtcars)
dim(mtcars)     # 32 11
# 32 rows, 11 columns
```
</details>

### The accessor: `$`

To grab a *single column* of a data frame, use `$`:

```r
mtcars$mpg          # the full mpg column as a vector
```

A crucial property: **the order of a column preserves the row order** — this is what lets you combine columns later.

```{r}
# YOUR CODE
# 1. mtcars$mpg — print the mpg column
# 2. mtcars$cyl
# 3. compute the mean of the mpg column with mean()
# 4. what is the max hp?  (max(mtcars$hp))
```

<details>
<summary><b>Reveal answer</b></summary>

```r
mtcars$mpg
mtcars$cyl
mean(mtcars$mpg)     # 20.09062
max(mtcars$hp)       # 335
```
</details>

### (Optional) the murders dataset from `dslabs`

The lecture's running example uses `dslabs::murders`. If you want to follow along, install and load it:

```{r}
# install.packages("dslabs")   # uncomment to install once
library(dslabs)
data("murders")
class(murders)         # data.frame
str(murders)           # 51 obs, 5 vars: state, abb, region (factor), population, total
head(murders)
names(murders)
murders$region         # a factor
```

---
## 6 · Final check — predict the output

Now the real test. For **each** line, write down what you expect, then run. This is exactly the "understand, don't just run" skill the course wants.

```{r}
# Q1
x <- 5
y <- "5"
class(x)
class(y)
x + 1          # works?
# y + 1        # what would this do? (uncomment after predicting)
```

<details>
<summary><b>Q1 answer</b></summary>

```r
class(x)    # "numeric"
class(y)    # "character" — "5" is a STRING, not a number
x + 1       # [1] 6  (numeric + numeric)
y + 1       # Error: non-numeric argument to binary operator  (can't add string + number)
```

The trap: `"5"` looks like 5 but is a **character**, and you can't do arithmetic on it. This is the difference between data that's *meant* to be a number and data that's *stored* as text.
</details>

```{r}
# Q2 — the vectorization question
v <- c(1, 2, 3, 4, 5)
v * 2
v + c(10, 20)
# what does R do when the lengths don't match? (predict, then run)
```

<details>
<summary><b>Q2 answer</b></summary>

```r
v * 2             # [1]  2  4  6  8 10   (elementwise multiply — no loop needed!)
v + c(10, 20)     # [1] 11 22 13 24 15   (RECYCLING: R recycles the shorter vector 10,20 → 10,20,10,20,10)
```

Two huge R ideas here:
- **Vectorization**: you operate on the *whole* vector at once, no `for` loop. This is why R is fast.
- **Recycling**: when lengths differ, R repeats the shorter one. Great when intended (b + 1), dangerous when not.

Challenge: `v + c(10, 20, 30)` — what happens? Try it. Does R warn you?
</details>

---
## Recap — what you should now be able to explain

1. Why R uses `<-` instead of `=`, and what each means.
2. Why calling `ls` alone (no parens) doesn't list objects.
3. The difference between `3`, `3L`, `"3"`, and `TRUE` — and what `class()` says about each.
4. Why `is.integer(3)` is `FALSE`.
5. R being 1-indexed vs Python being 0-indexed.
6. What a data frame is, and how `$`, `str()`, `head()`, `names()`, `dim()` help you inspect one.
7. Vectorization and recycling.

If you can answer these without looking, you've genuinely got Lec 1. Next: `lec02-tidyverse` → dplyr verbs and importing data.
